---
title: "Event alignment and AOI-linked biometric workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Event alignment and AOI-linked biometric workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

# Scope

This article shows a conservative workflow for aligning Gazepoint-derived biometric streams with task events, gaze records, and areas of interest (AOIs).

The goal is to make synchronization and feature-engineering decisions explicit before analysis: extract events, audit timing, align biometric streams to task markers, join gaze and biometric records, build AOI time courses, summarize AOI-linked biometric windows, and prepare model-ready tables.

The workflow is descriptive. Event-locked and AOI-linked biometric summaries are treated as synchronized measurement features. They are not interpreted here as direct evidence of attention, emotion, stress, cognitive load, clinical status, engagement, or psychological state.

# Workflow overview

A typical event-alignment and AOI-linked biometric workflow is:

1. Import or simulate gaze, event, and biometric streams.
2. Standardize column names and validate expected fields.
3. Extract TTL or task-event markers.
4. Audit timing, resets, sampling irregularity, and synchronization drift.
5. Align biometric streams to task events.
6. Synchronize gaze and biometric records.
7. Join biometric records to master trial metadata.
8. Build AOI time courses or AOI-window summaries.
9. Summarize event-locked multimodal features.
10. Prepare model-ready AOI and biometric tables.
11. Export decision logs and reproducibility records.

# Example data

For public documentation, use synthetic or example data rather than private Gazepoint exports.

```{r}
library(gpbiometrics)

bio <- simulate_gazepoint_biometrics(
  n_participants = 8,
  n_trials = 12,
  samples_per_trial = 120
)

eye <- simulate_gazepoint_eye_data(
  n_participants = 8,
  n_trials = 12,
  samples_per_trial = 120
)

master <- data.frame(
  participant_id = rep(sprintf("P%02d", 1:8), each = 12),
  trial_id = rep(seq_len(12), times = 8),
  condition = rep(c("A", "B"), length.out = 96)
)
```

# Step 1: standardize and validate streams

Start by standardizing names and checking that the expected identifiers and timing fields are available.

```{r}
bio_std <- standardise_gazepoint_biometric_names(bio)
eye_std <- standardize_gazepoint_column_names(eye)

validate_gazepoint_biometrics(bio_std)
validate_gazepoint_format(eye_std)
validate_gazepoint_metadata(master)
```

Timing checks should be performed before alignment. These checks help identify missing timestamps, irregular sampling, clock resets, and other synchronization risks.

```{r}
detect_gazepoint_biometric_timebase(bio_std)

assess_gazepoint_sampling_irregularity(
  bio_std,
  time_col = "TIME_MS",
  participant_col = "participant_id",
  trial_col = "trial_id"
)

assess_gazepoint_time_resets(
  bio_std,
  time_col = "TIME_MS",
  participant_col = "participant_id"
)
```

# Step 2: extract task events

Task events or TTL markers provide the timing anchors used for event-locked alignment. Event extraction should preserve the original timestamp and marker labels.

```{r}
events <- extract_gazepoint_ttl_events(
  bio_std,
  event_col = "EVENT",
  time_col = "TIME_MS",
  participant_col = "participant_id"
)

events
```

Event coverage should be checked before alignment. Missing or duplicated events can create ambiguous trial windows.

```{r}
audit_gazepoint_event_coverage(
  events,
  participant_col = "participant_id",
  event_col = "EVENT"
)
```

# Step 3: match events to biometric samples

Event matching links task markers to the nearest or enclosing biometric samples. The matching rule and tolerance should be reported.

```{r}
event_matches <- match_gazepoint_events_to_biometrics(
  events = events,
  biometrics = bio_std,
  event_time_col = "TIME_MS",
  biometric_time_col = "TIME_MS",
  participant_col = "participant_id",
  tolerance_ms = 100
)

event_matches
```

# Step 4: align biometric streams to events

Event alignment creates relative time fields such as milliseconds from stimulus onset, response onset, or trial start. These fields should be derived consistently across participants and trials.

```{r}
bio_event_locked <- align_gazepoint_biometrics_to_ttl(
  biometrics = bio_std,
  events = events,
  time_col = "TIME_MS",
  event_time_col = "TIME_MS",
  participant_col = "participant_id",
  event_col = "EVENT"
)

bio_event_locked
```

Multiple streams can also be aligned to shared event anchors.

```{r}
aligned_streams <- align_gazepoint_streams_by_events(
  streams = list(
    biometrics = bio_std,
    gaze = eye_std
  ),
  events = events,
  time_col = "TIME_MS",
  event_time_col = "TIME_MS",
  participant_col = "participant_id"
)
```

# Step 5: audit synchronization drift

Synchronization drift checks help identify whether timing offsets are stable across the recording. Drift should be reported as a data-quality property, not as a participant-level interpretation.

```{r}
drift <- audit_gazepoint_biometric_sync_drift(
  bio_event_locked,
  time_col = "TIME_MS",
  event_time_col = "event_time_ms",
  participant_col = "participant_id"
)

diagnose_gazepoint_sync_drift(drift)
```

# Step 6: synchronize biometrics with gaze

After event alignment, gaze and biometric records can be synchronized using common participant, trial, and relative-time fields.

```{r}
gaze_bio <- sync_gazepoint_biometrics_with_gaze(
  gaze = eye_std,
  biometrics = bio_event_locked,
  gaze_time_col = "time_ms",
  biometric_time_col = "TIME_MS",
  participant_col = "participant_id",
  trial_col = "trial_id"
)

gaze_bio
```

If the project also uses gp3tools outputs, keep this join explicit and auditable.

```{r}
join_gazepoint_biometrics_to_gp3tools(
  biometrics = bio_event_locked,
  gp3tools_data = eye_std,
  participant_col = "participant_id",
  trial_col = "trial_id"
)
```

# Step 7: join master trial metadata

Trial-level metadata should be joined after stream alignment so that condition labels, screen identifiers, and design variables are carried into downstream summaries.

```{r}
bio_master <- join_gazepoint_biometrics_to_master(
  biometrics = bio_event_locked,
  master = master,
  participant_col = "participant_id",
  trial_col = "trial_id"
)

bio_master
```

# Step 8: build AOI time courses

AOI time courses summarize when samples fall inside predefined regions. AOI membership should be treated as a screen-coordinate classification, not as direct evidence of attention or comprehension.

```{r}
aoi_definitions <- data.frame(
  aoi = c("header", "claim", "image", "button"),
  x_min = c(0, 100, 600, 700),
  x_max = c(1920, 550, 1300, 1000),
  y_min = c(0, 200, 200, 800),
  y_max = c(150, 700, 750, 950)
)

aoi_timecourse <- build_gazepoint_aoi_timecourse(
  gaze = eye_std,
  aoi = aoi_definitions,
  x_col = "gaze_x",
  y_col = "gaze_y",
  time_col = "time_ms",
  participant_col = "participant_id",
  trial_col = "trial_id"
)

aoi_timecourse
```

# Step 9: summarize AOI-linked biometrics

AOI-linked biometric summaries aggregate biometric samples over windows associated with AOI membership. These summaries describe synchronized signal values during AOI-defined intervals.

```{r}
aoi_bio <- summarise_gazepoint_aoi_biometrics(
  gaze_biometrics = gaze_bio,
  aoi_col = "aoi",
  participant_col = "participant_id",
  trial_col = "trial_id"
)

aoi_bio
```

For designs with predefined task windows, combine AOI summaries with event-locked windows.

```{r}
event_windows <- summarize_gazepoint_eventlocked_multimodal(
  bio_master,
  time_col = "time_from_event_ms",
  participant_col = "participant_id",
  trial_col = "trial_id",
  window = c(0, 3000)
)

multimodal_windows <- summarise_gazepoint_multimodal_windows(
  bio_master,
  participant_col = "participant_id",
  trial_col = "trial_id"
)
```

# Step 10: visualize multimodal timing

Timeline plots are useful for checking event alignment, AOI windows, and biometric summaries before modelling.

```{r}
plot_gazepoint_multimodal_timeline(
  bio_master,
  events = events,
  time_col = "TIME_MS",
  participant_col = "participant_id",
  trial_col = "trial_id"
)
```

# Step 11: prepare model-ready tables

Model-ready tables should contain identifiers, condition variables, AOI labels, event windows, synchronized biometric features, and QC indicators.

```{r}
aoi_model <- prepare_gazepoint_aoi_biometrics_model_data(
  aoi_bio,
  participant_col = "participant_id",
  trial_col = "trial_id",
  aoi_col = "aoi"
)

multimodal_model <- prepare_gazepoint_multimodal_model_data(
  multimodal_windows,
  participant_col = "participant_id",
  trial_col = "trial_id"
)

aoi_model
multimodal_model
```

# Reporting outputs

A reproducible event-alignment and AOI-linked biometric section should report timing anchors, event-matching tolerances, synchronization checks, AOI definitions, window definitions, feature summaries, and exclusion or QC rules.

```{r}
decision_log <- create_gazepoint_analysis_decision_log(
  decisions = data.frame(
    step = c(
      "event_extraction",
      "event_matching",
      "stream_alignment",
      "gaze_biometric_sync",
      "aoi_timecourse",
      "model_ready_tables"
    ),
    decision = c(
      "Task events were extracted before stream alignment.",
      "Events were matched to biometric samples using a documented tolerance.",
      "Biometric and gaze streams were aligned to shared timing anchors.",
      "Gaze and biometric records were synchronized before AOI summaries.",
      "AOI membership was computed from screen-coordinate rules.",
      "Model-ready tables retained identifiers, windows, features, and QC fields."
    )
  )
)

create_gazepoint_qc_supplement(
  qc_overview = drift,
  decision_log = decision_log
)

create_gazepoint_reproducibility_statement()
```

# Recommended reporting language

Use precise synchronization and feature-engineering language:

- Report event markers, timing anchors, and matching tolerances.
- Report sampling irregularity, time resets, and synchronization drift checks.
- Report AOI definitions as coordinate-based regions.
- Report AOI-linked biometric summaries as synchronized window-level features.
- Report the identifiers and QC fields retained in model-ready tables.
- Avoid describing AOI-linked biometric summaries as direct evidence of attention, emotion, stress, cognitive load, engagement, clinical state, or psychological response unless the study design and validation evidence explicitly support that interpretation.

# Minimal checklist

Before modelling event-locked or AOI-linked biometric outcomes, confirm that the analysis has:

- identified the original timing columns and event markers;
- checked sampling irregularity and time resets;
- documented event-matching tolerance and alignment rules;
- audited synchronization drift;
- defined AOIs using transparent coordinate rules;
- joined biometric, gaze, event, and trial metadata explicitly;
- summarized features by participant, trial, event window, and AOI where relevant;
- retained QC indicators in model-ready tables;
- kept a decision log for exclusions and transformations;
- used synthetic or anonymized examples in public documentation.

# Next steps

After event alignment and AOI-linked summaries, prepare reporting, reproducibility, and synthetic-data showcase materials so that reviewers can inspect how raw streams are transformed into analysis-ready tables.
